---
title: "Exo_suppl_beers"
author: "Kevin Missault"
date: "30/01/2019"
output: html_document
---


# Load packages

```{r loading packages, message = FALSE, warning = FALSE}
library(readr)
library(dplyr)
library(tidyr)
library(ggplot2)
library(lubridate)
```

# Read file

Read file and add delim option or use read_csv2() for ; as separator and , for the decimal point (common in European countries).
This table was created for the exercice. It's not real data. You can see the beer consumption (Liter) per year at different age according to working statue and gender.

```{r Beers_file}
beers_data <- read_delim("Data/Beers.csv", ";")
str(beers_data)
```

# Remove NA data

```{r Remove NA data, warning = FALSE}

is.na(beers_data)

beers_data_filter <- beers_data %>%
  filter(!is.na(Consumption))
beers_data_filter
```
# Mutate for date

Create a column Date using mutate from dplyr. What is the class of Date ?

```{r mutate_date}
beers_data_filter %>% 
  mutate(Date = paste(beers_data_filter$Year, beers_data_filter$Month, beers_data_filter$Day, sep = "-"))
```

# Lubridate for date

Create a column Date using lubridate. What is the class of Date ?
```{r lubridate for date}
beers_data_filter$Date <- ymd(paste(beers_data_filter$Year, beers_data_filter$Month, beers_data_filter$Day, sep = "-"))
str(beers_data_filter)
```

# Filter, select and arrange + export data

Export a table in a CSV file of women older than 35 years old, employed, select all columns except Day, Month and Year, and finally order in descending value of consumption of beers.

```{r Filter, select and arrange}
exported_table <- beers_data_filter %>% 
  filter(Age > 35) %>% 
  filter(Gender == "Female") %>%
  filter(Work == "Employed") %>% 
  select(-Age, -Day, -Month, -Year) %>% 
  arrange(desc(Consumption))

write_csv(exported_table, path = "Data_output/exported_table.csv")

read_csv("Data_output/exported_table.csv")
```

# Analysing data

Does working statue have an impact on beer consumption ?

```{r group_by work}
beers_data_filter %>% 
  group_by(Work) %>% 
  summarize(mean_consumption = mean(Consumption))
```

Do men drink more beers than women ?
```{r group_by gender}
beers_data_filter %>% 
  group_by(Gender) %>% 
  summarize(mean_consumption = mean(Consumption))
```


Does working statue have an influence on beer consumption according to gender ?
```{r group_by gender and work}
beers_data_filter %>% 
  group_by(Gender, Work) %>% 
  summarize(mean_consumption = mean(Consumption))
```

Do men drink more than women according to age and working statue ? Write code to get the following plot.

```{r ggplot2, echo = FALSE}
mean_age <- beers_data_filter %>%
  group_by(Gender, Age, Work) %>% 
  summarize(mean_consumption = mean(Consumption))

ggplot(data = mean_age, mapping = aes(x = Age, y = mean_consumption, color = Work)) +
  geom_line() +
  geom_point() +
  facet_wrap(~ Gender) +
  theme_bw() +
  scale_x_continuous(breaks = c(25, 35, 45, 55)) +
  labs(title = "Mean consumption of beers according to age, working statue and gender",
       x = "Age",
       y = "Mean consumption (Liter)")
```

```{r session_info}
sessionInfo()
```

