# Starting with data  {#sec:startdata}

**Learning Objectives**

- Describe what a data frame is.
- Load external data from a .csv file into a data frame.
- Summarize the contents of a data frame.
- Describe what a factor is.
- Convert between strings and factors.
- Reorder and rename factors.
- Change how character strings are handled in a data frame.
- Format dates.
- Export and save data.

## Presentation of the Survey Data

We are studying the species repartition and weight of animals caught
in plots in our study area. The dataset is stored as a comma separated
value (CSV) file.  Each row holds information for a single animal, and
the columns represent:

| Column           | Description                        |
|------------------|------------------------------------|
| record\_id       | Unique id for the observation      |
| month            | month of observation               |
| day              | day of observation                 |
| year             | year of observation                |
| plot\_id         | ID of a particular plot            |
| species\_id      | 2-letter code                      |
| sex              | sex of animal ("M", "F")           |
| hindfoot\_length | length of the hindfoot in mm       |
| weight           | weight of the animal in grams      |
| genus            | genus of animal                    |
| species          | species of animal                  |
| taxon             | e.g. Rodent, Reptile, Bird, Rabbit |
| plot\_type       | type of plot                       |

We are going to use the R function `download.file()` to download the
CSV file that contains the survey data from figshare, and we will use
`read.csv()` to load into memory the content of the CSV file as an
object of class `data.frame`.  Inside the download.file command, the
first entry is a character string with the source URL
("https://ndownloader.figshare.com/files/2292169"). This source URL
downloads a CSV file from figshare. The text after the comma
("data/portal_data_joined.csv") is the destination of the file on your
local machine. You'll need to have a folder on your machine called
"data" where you'll download the file. So this command downloads a
file from figshare, names it "portal_data_joined.csv," and adds it to
a preexisting folder named "data."

```{r, eval=FALSE, purl=TRUE}
download.file(url="https://ndownloader.figshare.com/files/2292169",
              destfile = "data/portal_data_joined.csv")
```

You are now ready to load the data:

```{r, eval=TRUE, purl=FALSE}
surveys <- read.csv("data/portal_data_joined.csv")
```

This statement doesn't produce any output because, as you might
recall, assignments don't display anything. If we want to check that
our data has been loaded, we can see the contents of the data frame by
typing its name: `surveys`.

Wow... that was a lot of output. At least it means the data loaded
properly. Let's check the top (the first 6 lines) of this data frame
using the function `head()`:

```{r, results='show', purl=FALSE}
head(surveys)
## Try also
## View(surveys)
```

**Note**

`read.csv` assumes that fields are delineated by commas, however, in
several countries, the comma is used as a decimal separator and the
semicolon (;) is used as a field delineator. If you want to read in
this type of files in R, you can use the `read.csv2` function. It
behaves exactly like `read.csv` but uses different parameters for the
decimal and the field separators. If you are working with another
format, they can be both specified by the user. Check out the help for
`read.csv()` by typing `?read.csv` to learn more. There is also the
`read.delim()` for in tab separated data files. It is important to
note that all of these functions are actually wrapper functions for
the main `read.table()` function with different arguments.  As such,
the surveys data above could have also been loaded by using
`read.table()` with the separation argument as `,`. The code is as
follows:

```
surveys <- read.table(file="data/portal_data_joined.csv", sep=",", header=TRUE)
```

The header argument has to be set to TRUE to be able to read the
headers as by default `read.table()` has the header argument set to
FALSE.

## What are data frames?

Data frames are the _de facto_ data structure for most tabular data,
and what we use for statistics and plotting.

A data frame can be created by hand, but most commonly they are generated by the
functions `read.csv()` or `read.table()`; in other words, when importing
spreadsheets from your hard drive (or the web).

A data frame is the representation of data in the format of a table where the
columns are vectors that all have the same length. Because columns are
vectors, each column must contain a single type of data (e.g., characters, integers,
factors). For example, here is a figure depicting a data frame comprising a
numeric, a character, and a logical vector.

![](./figs/data-frame.svg)


We can see this when inspecting the <b>str</b>ucture of a data frame
with the function `str()`:

```{r, purl=FALSE}
str(surveys)
```

## Inspecting `data.frame` Objects

We already saw how the functions `head()` and `str()` can be useful to check the
content and the structure of a data frame. Here is a non-exhaustive list of
functions to get a sense of the content/structure of the data. Let's try them out!

**Size**:

- `dim(surveys)` - returns a vector with the number of rows in the
   first element, and the number of columns as the second element (the
   **dim**ensions of the object)
- `nrow(surveys)` - returns the number of rows
- `ncol(surveys)` - returns the number of columns

**Content**:

- `head(surveys)` - shows the first 6 rows
- `tail(surveys)` - shows the last 6 rows

**Names**:

- `names(surveys)` - returns the column names (synonym of `colnames()`
   for `data.frame` objects)
- `rownames(surveys)` - returns the row names

**Summary**:

- `str(surveys)` - structure of the object and information about the
  class, length and content of each column
- `summary(surveys)` - summary statistics for each column


Note: most of these functions are "generic", they can be used on other types of
objects besides `data.frame`.


`r msmbstyle::question_begin()`

Based on the output of `str(surveys)`, can you answer the following
questions?

- What is the class of the object `surveys`?
- How many rows and how many columns are in this object?
- How many species (as defined by the `species_id` variable) have been
  recorded during these surveys?

`r msmbstyle::question_end()`


`r msmbstyle::solution_begin()`
- class: data frame
- how many rows: 34786,  how many columns: 13
- how many species: 48
`r msmbstyle::solution_end()`


## Indexing and subsetting data frames

Our survey data frame has rows and columns (it has 2 dimensions), if
we want to extract some specific data from it, we need to specify the
"coordinates" we want from it. Row numbers come first, followed by
column numbers. However, note that different ways of specifying these
coordinates lead to results with different classes.


```{r, eval = FALSE}
# first element in the first column of the data frame (as a vector)
surveys[1, 1]
# first element in the 6th column (as a vector)
surveys[1, 6]
# first column of the data frame (as a vector)
surveys[, 1]
# first column of the data frame (as a data.frame)
surveys[1]
# first three elements in the 7th column (as a vector)
surveys[1:3, 7]
# the 3rd row of the data frame (as a data.frame)
surveys[3, ]
# equivalent to head_surveys <- head(surveys)
head_surveys <- surveys[1:6, ]
```

`:` is a special function that creates numeric vectors of integers in
increasing or decreasing order, test `1:10` and `10:1` for
instance. See section \@ref(sec:genvec) for details.

You can also exclude certain indices of a data frame using the "`-`" sign:

```{r, eval = FALSE}
surveys[, -1]          ## The whole data frame, except the first column
surveys[-c(7:34786), ] ## Equivalent to head(surveys)
```

Data frames can be subset by calling indices (as shown previously),
but also by calling their column names directly:

```{r, eval = FALSE, purl=FALSE}
surveys["species_id"]       # Result is a data.frame
surveys[, "species_id"]     # Result is a vector
surveys[["species_id"]]     # Result is a vector
surveys$species_id          # Result is a vector
```

In RStudio, you can use the autocompletion feature to get the full and
correct names of the columns.

`r msmbstyle::question_begin()`

1. Create a `data.frame` (`surveys_200`) containing only the data in
   row 200 of the `surveys` dataset.

2. Notice how `nrow()` gave you the number of rows in a `data.frame`?

- Use that number to pull out just that last row in the inital
  `surveys` data frame.

- Compare that with what you see as the last row using `tail()` to make
  sure it's meeting expectations.

- Pull out that last row using `nrow()` instead of the row number.

- Create a new data frame (`surveys_last`) from that last row.

3. Use `nrow()` to extract the row that is in the middle of the
   `surveys` dataframe. Store the content of this row in an object
   named `surveys_middle`.

4. Combine `nrow()` with the `-` notation above to reproduce the
   behavior of `head(surveys)`, keeping just the first through 6th
   rows of the surveys dataset.

`r msmbstyle::question_end()`

`r msmbstyle::solution_begin()`
> ```{r}
> ## 1.
> surveys_200 <- surveys[200, ]
> ## 2.
> ## Saving `n_rows` to improve readability and reduce duplication
> n_rows <- nrow(surveys)
> surveys_last <- surveys[n_rows, ]
> ## 3.
> surveys_middle <- surveys[n_rows / 2, ]
> ## 4.
> surveys_head <- surveys[-(7:n_rows), ]
> ```
`r msmbstyle::solution_end()`

## Factors


When we did `str(surveys)` we saw that several of the columns consist
of integers. The columns `genus`, `species`, `sex`, `plot_type`,
... however, are of a special class called `factor`. Factors are very
useful and actually contribute to making R particularly well suited to
working with data. So we are going to spend a little time introducing
them.

Factors represent **categorical data**. They are stored as integers
associated with labels and they can be ordered or unordered. While
factors look (and often behave) like character vectors, they are
actually treated as integer vectors by R. So you need to be very
careful when treating them as strings.

Once created, factors can only contain a pre-defined set of values,
known as *levels*. By default, R always sorts levels in alphabetical
order. For instance, if you have a factor with 2 levels:

```{r, purl=TRUE}
sex <- factor(c("male", "female", "female", "male"))
```

R will assign `1` to the level `"female"` and `2` to the level
`"male"` (because `f` comes before `m`, even though the first element
in this vector is `"male"`). You can see this by using the function
`levels()` and you can find the number of levels using `nlevels()`:


```{r, purl=FALSE}
levels(sex)
nlevels(sex)
```

Sometimes, the order of the factors does not matter, other times you
might want to specify the order because it is meaningful (e.g., "low",
"medium", "high"), it improves your visualization, or it is required
by a particular type of analysis. Here, one way to reorder our levels
in the `sex` vector would be:


```{r}
sex ## current order
sex <- factor(sex, levels = c("male", "female"))
sex ## after re-ordering
```

In R's memory, these factors are represented by integers (1, 2, 3),
but are more informative than integers because factors are self
describing: `"female"`, `"male"` is more descriptive than `1`,
`2`. Which one is "male"?  You wouldn't be able to tell just from the
integer data. Factors, on the other hand, have this information built
in. It is particularly helpful when there are many levels (like the
species names in our example dataset).

### Converting to factors

If you need to convert a factor to a character vector, you use
`as.character(x)`.

```{r, purl=FALSE}
as.character(sex)
```

<!-- ### Numeric factors {-} -->

<!-- Converting factors where the levels appear as numbers (such as -->
<!-- concentration levels, or years) to a numeric vector is a little -->
<!-- trickier. The `as.numeric()` function returns the index values of the -->
<!-- factor, not its levels, so it will result in an entirely new (and -->
<!-- unwanted in this case) set of numbers.  One method to avoid this is to -->
<!-- convert factors to characters, and then to numbers.  Another method is -->
<!-- to use the `levels()` function. Compare: -->

<!-- ```{r} -->
<!-- year_fct <- factor(c(1990, 1983, 1977, 1998, 1990)) -->
<!-- as.numeric(year_fct)  ## Wrong! And there is no warning... -->
<!-- as.numeric(as.character(year_fct)) ## Works... -->
<!-- as.numeric(levels(year_fct))[year_fct] ## The recommended way. -->
<!-- ```

<!-- Notice that in the `levels()` approach, three important steps occur: -->

<!-- * We obtain all the factor levels using `levels(year_fct)` -->
<!-- * We convert these levels to numeric values using `as.numeric(levels(year_fct))` -->
<!-- * We then access these numeric values using the underlying integers of the -->
<!--   vector `year_fct` inside the square brackets -->

### Renaming factors

When your data is stored as a factor, you can use the `plot()`
function to get a quick glance at the number of observations
represented by each factor level. Let's look at the number of males
and females captured over the course of the experiment:

```{r firstfactorplot, fig.cap = "Bar plot of the number of females and males captured during the experiment."}
plot(surveys$sex)
```

In addition to males and females, there are about 1700 individuals for
which the sex information hasn't been recorded. Additionally, for
these individuals, there is no label to indicate that the information
is missing or undetermined. Let's rename this label to something more
meaningful. Before doing that, we're going to pull out the data on sex
and work with that data, so we're not modifying the working copy of
the data frame:

```{r, results=TRUE, purl=FALSE}
sex <- surveys$sex
head(sex)
levels(sex)
levels(sex)[1] <- "undetermined"
levels(sex)
head(sex)
```

`r msmbstyle::question_begin()`

- Rename "F" and "M" to "female" and "male" respectively.

- Now that we have renamed the factor level to "undetermined", can you
  recreate the barplot such that "undetermined" is last (after
  "male")?

`r msmbstyle::question_end()`

`r msmbstyle::solution_begin()`

```{r}
levels(sex)[2:3] <- c("female", "male")
sex <- factor(sex, levels = c("female", "male", "undetermined"))
plot(sex)
```

`r msmbstyle::solution_end()`


### Using `stringsAsFactors=FALSE`

By default, when building or importing a data frame, the columns that
contain characters (i.e. text) are coerced (= converted) into
factors. Depending on what you want to do with the data, you may want
to keep these columns as `character`. To do so, `read.csv()` and
`read.table()` have an argument called `stringsAsFactors` which can be
set to `FALSE`.

In most cases, it is preferable to set `stringsAsFactors = FALSE` when
importing data and to convert as a factor only the columns that
require this data type.


```{r, eval=FALSE, purl=TRUE}
## Compare the difference between our data read as `factor` vs `character`.
surveys <- read.csv("data/portal_data_joined.csv", stringsAsFactors = TRUE)
str(surveys)
surveys <- read.csv("data/portal_data_joined.csv", stringsAsFactors = FALSE)
str(surveys)
## Convert the column "plot_type" into a factor
surveys$plot_type <- factor(surveys$plot_type)
```


`r msmbstyle::question_begin()`
We have seen how data frames are created when using `read.csv()`, but
they can also be created by hand with the `data.frame()` function.
There are a few mistakes in this hand-crafted `data.frame`. Can you
spot and fix them?  Don't hesitate to experiment!

```{r, eval=FALSE, purl=FALSE}
animal_data <- data.frame(
       animal = c(dog, cat, sea cucumber, sea urchin),
       feel = c("furry", "squishy", "spiny"),
       weight = c(45, 8 1.1, 0.8))
```
`r msmbstyle::question_end()`


`r msmbstyle::solution_begin()`
- missing quotations around the names of the animals
- missing one entry in the "feel" column (probably for one of the furry animals)
- missing one comma in the weight column
`r msmbstyle::solution_end()`


`r msmbstyle::question_begin()`

Can you predict the class for each of the columns in the following
example?

Check your guesses using `str(country_climate)`:

- Are they what you expected?  Why? Why not?

- What would have been different if we had added `stringsAsFactors =
  FALSE` when creating the data frame?

- What would you need to change to ensure that each column had the
  accurate data type?

```{r, eval=FALSE, purl=FALSE}
country_climate <- data.frame(
       country = c("Canada", "Panama", "South Africa", "Australia"),
       climate = c("cold", "hot", "temperate", "hot/temperate"),
       temperature = c(10, 30, 18, "15"),
       northern_hemisphere = c(TRUE, TRUE, FALSE, "FALSE"),
       has_kangaroo = c(FALSE, FALSE, FALSE, 1)
       )
```
`r msmbstyle::question_end()`


`r msmbstyle::solution_begin()`
- `country`, `climate`, `temperature`, and `northern_hemisphere` are
  factors; `has_kangaroo` is numeric.

- Using `stringsAsFactors = FALSE` would have made them character
  instead of factors.

- Removing the quotes in temperature and northern_hemisphere and
  replacing 1 by TRUE in the `has_kangaroo` column would give what was
  probably intended.

`r msmbstyle::solution_end()`

The automatic conversion of data type is sometimes a blessing, sometimes an
annoyance. Be aware that it exists, learn the rules, and double check that data
you import in R are of the correct type within your data frame. If not, use it
to your advantage to detect mistakes that might have been introduced during data
entry (a letter in a column that should only contain numbers for instance).

Learn more in this [RStudio
tutorial](https://support.rstudio.com/hc/en-us/articles/218611977-Importing-Data-with-RStudio)

## Matrices

Before proceeding, now that we have learnt about dataframes, let's
recap package installation and learn about a new data type, namely the
`matrix`. Like a `data.frame`, a matrix has two dimensions, rows and
columns. But the major difference is that all cells in a `matrix` must
be of the same type: `numeric`, `character`, `logical`, ... In that
respect, matrices are closer to a `vector` than a `data.frame`.

Let's generate a matrix containing information about all packages that
are currently installed on our respective computers using the
`installed.packages()` function. We will use the same function to
explore this matrix as those used for dataframes above.

```{r ip}
ip <- installed.packages()
dim(ip)
head(colnames(ip))
head(rownames(ip)) ## these are the package names
## View(ip)
```

```{r pkg_qst, purl=FALSE, echo=FALSE}
msmbstyle::question(text = "Using the function `installed.packages()`, verify if the `lubridate` package is installed. If not, install it from CRAN.")
```

`r msmbstyle::solution_begin()`

```{r pkg_sln, purl=FALSE, eval=FALSE}
## 1. get the names of all installed packages
ip <- rownames(installed.packages())
## 2. TRUE if installed, FALSE otherwise
"lubridate" %in% ip
## 3. install the package
install.package("lubridate")
```

`r msmbstyle::solution_end()`


The default constructor for a matrix is `matrix`. It takes a vector of
values to populate the matrix and the number of row and/or
columns[^ncol]. The values are sorted along the columns, as illustrated
below.

```{r mat1}
m <- matrix(1:9, ncol = 3, nrow = 3)
m
```

[^ncol]: Either the number of rows or columns are enough, as the other
one can be deduced from the length of the values. Try out what happens
if the values and number of rows/columns don't add up.


It is often useful to create large random data matrices as test
data. The exercise below asks you to create such a matrix with random
data drawn from a normal distribution of mean 0 and standard deviation
1, which can be done with the `rnorm()` function.

```{r rnormmat_qst, purl=FALSE, echo=FALSE}
msmbstyle::question(text = "Construct a matrix of dimension 1000 by 3 of normally distributed data (mean 0, standard deviation 1).")
```


`r msmbstyle::solution_begin()`

```{r rnormmat_sln, purl=FALSE}
set.seed(123)
m <- matrix(rnorm(3000), ncol = 3)
dim(m)
head(m)
```

`r msmbstyle::solution_end()`


## Formatting Dates

One of the most common issues that new (and experienced!) R users have is
converting date and time information into a variable that is appropriate and
usable during analyses. As a reminder from earlier in this lesson, the best
practice for dealing with date data is to ensure that each component of your
date is stored as a separate variable. Using `str()`, We can confirm that our
data frame has a separate column for day, month, and year, and that each contains
integer values.

```{r, eval=FALSE, purl=FALSE}
str(surveys)
```

We are going to use
the `ymd()` function from the package **`lubridate`** (which belongs
to the **`tidyverse`**; learn more
[here](https://www.tidyverse.org/)). . **`lubridate`** gets installed
as part as the **`tidyverse`** installation. When you load the
**`tidyverse`** (`library(tidyverse)`), the core packages (the
packages used in most data analyses) get loaded. **`lubridate`**
however does not belong to the core tidyverse, so you have to load it
explicitly with `library(lubridate)`

Start by loading the required package:

```{r loadlibridate, message=FALSE, purl=FALSE}
library("lubridate")
```

`ymd()` takes a vector representing year, month, and day, and converts it to a
`Date` vector. `Date` is a class of data recognized by R as being a date and can
be manipulated as such. The argument that the function requires is flexible,
but, as a best practice, is a character vector formatted as "YYYY-MM-DD".


Let's create a date object and inspect the structure:

```{r}
my_date <- ymd("2015-01-01")
str(my_date)
```

Now let's paste the year, month, and day separately - we get the same result:

```{r}
# sep indicates the character to use to separate each component
my_date <- ymd(paste("2015", "1", "1", sep = "-"))
str(my_date)
```

Now we apply this function to the surveys dataset. Create a character
vector from the `year`, `month`, and `day` columns of `surveys` using
`paste()`:

```{r}
head(paste(surveys$year, surveys$month, surveys$day, sep = "-"))
```

This character vector can be used as the argument for `ymd()`:

```{r}
head(ymd(paste(surveys$year, surveys$month, surveys$day, sep = "-")))
```

The resulting `Date` vector can be added to `surveys` as a new column called `date`:

```{r, purl=FALSE}
surveys$date <- ymd(paste(surveys$year, surveys$month, surveys$day, sep = "-"))
str(surveys) # notice the new column, with 'date' as the class
```

Let's make sure everything worked correctly. One way to inspect the
new column is to use `summary()`:

```{r, results=TRUE, purl=FALSE}
summary(surveys$date)
```

Something went wrong: some dates have missing values. Let's
investigate where they are coming from.

We can use the functions we saw previously to deal with missing data
to identify the rows in our data frame that are failing. If we combine
them with what we learned about subsetting data frames earlier, we can
extract the columns "year, "month", "day" from the records that have
`NA` in our new column `date`. We will also use `head()` so we don't
clutter the output:

```{r, results=TRUE, purl=FALSE}
is_missing_date <- is.na(surveys$date)
date_columns <- c("year", "month", "day")
missing_dates <- surveys[is_missing_date,  date_columns]

head(missing_dates)
```

Why did these dates fail to parse? If you had to use these data for your
analyses, how would you deal with this situation?

## Summary of R objects

So far, we have seen several types of R object varying in the number
of dimensions and whether they could store a single of multiple data
types:

- **`vector`**: one dimension (they have a length), single type of data.
- **`matrix`**: two dimensions, single type of data.
- **`data.frame`**: two dimensions, one type per column.

## Lists

A data type that we haven't seen yet, but that is useful to know, and
follows from the summary that we have just seen are lists:

- **`list`**: one dimension, every item can be of a different data
    type.

Below, let's create a list containing a vector of numbers, characters,
a matrix, a dataframe and another list:

```{r list0}
l <- list(1:10, ## numeric
          letters, ## character
          installed.packages(), ## a matric
          cars, ## a data.frame
          list(1, 2, 3)) ## a list
length(l)
str(l)
```

List subsetting is done using `[]` to subset a new sub-list or `[[]]`
to extract a single element of that list (using indices or names, of
the list is named).

```{r}
l[[1]] ## first element
l[1:2] ## a list of length 2
l[1]   ## a list of length 1
```
## Exporting and saving data {#sec:exportandsave}

### Exporting tabular data {-}

We have seen how to read a text-based spreadsheet into R using the
`read.table` family of functions. To export a `data.frame` to a
text-based spreadsheet, we can use the `write.table` set of functions
(`write.csv`, `write.delim`, ...). They all take the variable to be
exported and the file to be exported to. For example, to export the
`surveys` data to the `my_surveys.csv` file in the `data_output`
directory, we would execute:

```{r, eval = FALSE}
write.csv(surveys, file = "data_output/my_surveys.csv")
```
This new csv file can now be shared with other collaborators who
aren't familiar with R.

### Saving data {-}

Exporting data to a spreadsheet has several limitations, such as those
described in the first chapter such as possible inconsistencies with
`,` and `.` for decimal separators and lack of variable type
definitions. Furthermore, exporting data to a spreadsheet is only
relevant for rectangular data such as dataframes and matrices.

A more general way to save data, that is specific to R and is
guaranteed to work on any operating system, is to use the `save`
function. Saving objects will generate a binary representation of the
object on disk, a *R Data* file (`rda` extension) that guarantees to
produce the same object once loaded back into R using the `load`
function.

```{r, eval = FALSE}
save(surveys, file = "data_output/surveys.rda")
rm(surveys)
load("data_output/surveys.rda")
head(surveys)
```

Note about how the function `load` loads the object in the file
directly in the global environment.


There is also the `saveRDS` and `readRDS` functions that save R
objects to binary files (using the `rds` extension here) and read
these back into R. From a user's perspective, main different is that,
`load` loads an object in the gloval environment while `readRDS` reads
the data from disk and returns it. It is this necessary to store the
output of `readRDS` in a variable:

```{r}
saveRDS(surveys, file = "data_output/surveys.rds")
rm(surveys)
surveys <- readRDS("data_output/surveys.rds")
head(surveys)
```

```{r echo=FALSE}
unlink("data_output/surveys.rda")
unlink("data_output/surveys.rds")
```

To conclude, when it comes to saving data from R that will be loaded
again in R, saving and loading is the preferred approach. If tabular
data need to be shared with somebody that is not using R, then
exporting to a text-based spreadsheet is a good alternative.

## Additional exercises

`r msmbstyle::question_begin()`

You're doing an colony counting experiment, counting every day how
many molds you see in your cell cultures.

- Create a vector named `molds` containing the results of your counts:
  1, 2, 5, 8 and 10. Create a vector `days` containing the week day,
  from Monday to Friday. Use these two vector to create a `data.frame`
  named `molds_study` containing two variables, `Day` and
  `Molds_count`.

- Create a new `data.frame` that contains the observations where more
  that 2 colonies were counted. How many observations are there? How
  many counts are there in total for these observations.

You repeat the molds study experiment the following week and count the
following numbers of molds: 1, 6, 6, 5 and 4.

- Add these data as a third column to the `molds_study` `data.frame`
  and rename the variables as `Day`, `Molds_1` and `Molds_2`.

- Calculate for each experiment the total number of molds
  counted. Check if the first experiment counted more molds than the
  second one.

- Save the `molds_study` variable in a file named `molds_study.rda`.

`r msmbstyle::question_end()`

<!-- `r msmbstyle::solution_begin()` -->
<!-- ```{r} -->
<!-- molds <- c(1, 2, 5, 8, 10) -->
<!-- days <-  c("Monday", "Tuesday", "Wednesday", "Thursday", "Friday") -->
<!-- molds_study <- data.frame("Day" = days, "Molds_count" = molds) -->
<!-- ``` -->

<!-- ```{r} -->
<!-- molds2 <- molds_study[molds_study$Molds_count > 2, ] -->
<!-- nrow(molds2) -->
<!-- sum(molds2$Molds_count) -->
<!-- ``` -->

<!-- ```{r} -->
<!-- molds_study$Molds_count2 <- c(1, 6, 6, 5, 4) -->
<!-- names(molds_study)[2:3] <- c("Molds_1", "Molds_2") -->
<!-- ``` -->

<!-- ```{r} -->
<!-- total1 <- sum(molds_study$Molds_1) -->
<!-- total2 <- sum(molds_study$Molds_2) -->
<!-- total1 > total2 -->
<!-- ``` -->
<!-- `r msmbstyle::solution_end()` -->


`r msmbstyle::question_begin()`

We are going to analyse beer consumption in 48 individuals. The data
are available in the `rWSBIM1207` package. The data illustrated the
fictive beer consumption in liters per year at different age according
to gender and employment.

- Load the `rWSBIM1207` package. If the package isn't installed of its
  version is greater than 0.1.1, install it from the
  `UCLouvain-CBIO/rWSBIM1207` GitHub repository using the
  `devtools::install_github` function.

- Using the `beers.csv()` function from `rWSBIM1207`, find the path
  the `beers.csv` file and read it to produce a `data.frame` named
  `beers`. The spreadsheet uses semi-colons `;` to separate cells. Use
  `read.csv2` and `read.delim` and set the separator appropriately,
  and verify that the two variables are identical.

- Check the number of observations and identify the variables that are
  available. Calculate a summary of each variable using the `summary`
  function directly on the `data.frame`.

- Calculate the mean and the median age and consumption.

- Do men consume more beer than women on average? To answer this
  question, calculate the mean consumption for men only, selecting the
  observations with `Gender` equal to `Male`. Then do the same for
  observations with `Gender` equal to `Female`.

- Calculate a two-way table of gender and employment status.

- Remove observations with missing values and export the data into a
  new `csv` file called `beers_no_na.csv`.

`r msmbstyle::question_end()`


`r msmbstyle::question_begin()`

We are going to analyse clinical data from The Cancer Genome Atlas
(TCGA). The data are available in the `rWSBIM1207` package.

- Load the `rWSBIM1207` package. If the package isn't installed of its
  version is greater than 0.1.1, install it from the
  `UCLouvain-CBIO/rWSBIM1207` GitHub repository using the
  `devtools::install_github` function.

- Obtain the path to th csv file containing the clinical data need for
  this exercise using the `clinical1.csv` function and read it into R
  as a `data.frame` called `clinical`.

- Inspect the data using `str` and `View`. How many patients are
  recorded in the table?

- Print the column names using two different functions.

- Create a smaller data frame called `clinical_mini` containing only
  the columns corresponding to the `patientID`, `gender`,
  `age_at_diagnosis` and `smoking_history`. Try to do this using
  column indices and column names.

- Inspect the `smoking_history` column. How many categories are
  recorded? How many observations are there for each category?

- The column age at diagnosis is recorded in days. Create a new column
  `years_at_diagnosis` corresponding to the age at diagnosis converted
  in years.

- Calculate the mean and median age at diagnosis. Hint: pay attention
  to missing values!

- Is there a difference between the `years_at_diagnosis` for male and
  female patients?

- Use the `quantile` function to calculate the first and last quartile
  of age at diagnosis. Use the help function (`?quantile`) to see how
  to use the `quantile` function.

- Use the `summary` function to confirm your previous results.

`r msmbstyle::question_end()`


<!-- `r msmbstyle::solution_begin()` -->

<!-- ```{r} -->
<!-- library("rWSBIM1207") -->
<!-- clinical <- read.csv(clinical1.csv()) -->

<!-- str(clinical) -->
<!-- names(clinical) -->
<!-- ## or -->
<!-- colnames(clinical) -->

<!-- clinical_mini <- clinical[, c(1, 3, 4, 12)] -->
<!-- ## or -->
<!-- clinical_mini <- clinical[, c("patientID", "gender", "age_at_diagnosis", "smoking_history")] -->

<!-- levels(clinical_mini$smoking_history) -->
<!-- table(clinical_mini$smoking_history) -->

<!-- clinical_mini$years_at_diagnosis <- clinical_mini$age_at_diagnosis / 365 -->

<!-- mean(clinical_mini$years_at_diagnosis, na.rm = TRUE) -->
<!-- median(clinical_mini$years_at_diagnosis, na.rm = TRUE) -->

<!-- quantile(clinical_mini$years_at_diagnosis, 0.25, na.rm = TRUE) -->
<!-- quantile(clinical_mini$years_at_diagnosis, 0.75, na.rm = TRUE) -->

<!-- summary(clinical_mini$years_at_diagnosis) -->
<!-- ``` -->

<!-- `r msmbstyle::solution_end()` -->
