# Additional programming concepts

**Learning Objectives**

Learn programming concepts, including

- how to handle conditions
- iterate of data structures
- good coding practice
- code re-use through functions

When the size and the complexity of the data increases, or the data
science question of interest becomes more complex, the data analysis
techniques as we have seen them so far need to be complemented with
programming techniques. From a data science point of view, there is no
clear delimitation between data analysis and (data) programming, both
morphing into each other[^dataprog].

[^dataprog]: A fundamental difference however is how data analysis and
    programming are taught. When it comes to researchers, and
    biomedical researchers in particular, teaching programming to
    analyse data isn't successful. Teaching data analysis to
    eventually programme with data, however, has proven a successful
    strategy.

This chapter will introduce some additional programming skills and
demonstrate how to use them in the context of high troughput omics
data.

## Conditionals

`if` and `else`.

## Iteration

Iteration describes the situtation when an specific action has to be
repeated many times on different inputs of the same type. For example,
if we have a vector of numeric `x` shown below,

```{r}
(x <- 1:10)
```


`for` loops and `apply` functions.

Example: iteration over files.

Possibly also mention `while`.

## Writing clean code

Writing clean code means writing easily readable code, hence easily
understable code and, eventually code with less bugs.

Defensive programming


## Writing new functions

A function is composed of a name, inputs (inside the parenthesis), a
body (between curly brackets) and an ouput (last statement or variably
indise the `return` statement).

```{r myfun}
my_fun <- function(x, y) {
    message("First input: ", x)
    message("Second input: ", y)
    z <- x * abs(y)
    return(z)
}
my_fun(2, -5)
```
