# Additional programming concepts

**Learning Objectives**

Learn programming concepts, including

- how to handle conditions
- iterate of data structures
- good coding practice
- code re-use through functions

When the size and the complexity of the data increases, or the data
science question of interest becomes more complex, the data analysis
techniques as we have seen them so far need to be complemented with
programming techniques. From a data science point of view, there is no
clear delimitation between data analysis and (data) programming, both
morphing into each other[^dataprog].

[^dataprog]: A fundamental difference however is how data analysis and
    programming are taught. When it comes to researchers, and
    biomedical researchers in particular, teaching programming to
    analyse data isn't successful. Teaching data analysis to
    eventually programme with data, however, has proven a successful
    strategy.

This chapter will introduce some additional programming skills and
demonstrate how to use them in the context of high troughput omics
data.

## Writing clean code


Writing clean code means writing easily readable code, hence easily
understable code and, eventually code with less bugs. One easy way to
achieve this is through consistency, i.e. stick to a **style
guide**. This issue is that there are several style guides available,
often with conflicting suggestions. Two widely used ones are the
[Bioconductor style
guide](http://bioconductor.org/developers/how-to/coding-style/) and
[Hadley Wickhams's R Style
Guide](http://r-pkgs.had.co.nz/style.html). The advice from for naming
variables seen in the first and second chapters are also relevant.

Here are a couple of suggestions:

- Use `<-` to assign variables. Use `=` is also valid, but make sure
  that you stick with one.
- Use either camel case (`camelCaseNaming`) of snake case
  (`snake_case_naming`), and avoid using dots (don't use
  `dot.variable.names`). These conventions apply to functions,
  variables and files (for the latter, a `-` instead of `_` is also
  acceptable).
- Always spell out `TRUE` and `FALSE`, and resist the temptation to
  use `T` and `F` instead.
- Use 4 spaces for indenting. No tabs.
- No lines longer than 80 characters.
- Use spaces around binary operators.
- No spaces between a function name and the opening parenthesis.

Another good advise is to avoid re-writing the same code many
times. We will see below two strategies to do that, namely iteration
and writing new functions. This firstly reduces the amount of code
typed and hence the number of bugs, but more importantly enables
consistency. If something in your code changes, there's only one
change and it applies everywhere, rather than doing that same change
repeatedly (at the first of adding bugs and to miss some updates).

When writing code, **keep it simple and short**[^KISS]. Ideally, the
code should be evident. But when the questions tackled are not
trivial, is becomes essential to add **comments** to clarify aspects
of the script/programme. Make sure to use them to describe why
something is done, rather than explaining how things are done (which
is tyically best done by the code itself).

[^KISS]: The official motto is
[KISS](https://en.wikipedia.org/wiki/KISS_principle), *keep your
functions simple, stupid*, widely used in programming.

A general guideline to avoid bugs is to apply **defensive
programming**:

- making the code work in a predicable manner

- writing code that fails in a well-defined manner

- if something weird happens, either properly deal with it, of fail
  quickly and loudly




## Writing new functions

A function is composed of a name, inputs (inside the parenthesis), a
body (between curly brackets) and an ouput (last statement or variably
indise the `return` statement).

```{r myfun}
my_fun <- function(x, y) {
    message("First input: ", x)
    message("Second input: ", y)
    z <- x * abs(y)
    return(z)
}
my_fun(2, -5)
```

## Conditionals

`if` and `else`.

## Iteration

Iteration describes the situtation when a specific operation has to be
repeated many times on different inputs of the same type. For example,
if we have a vector of numeric `x` shown below,

```{r}
(x <- 1:10)
```

and we wanted to apply the logarithm operation on each element of `x`,
it wouldn't be convenient to type

```
log(1)
log(2)
log(3)
...
log(10)
```

The concept of iteration allows us to programme the following command:

> Repeat `log` for each value of my input `x`.


or more formally

> Repeat `log(i)` where `i` takes in turn each value of my input `x`.


We will see different ways of implementing such an iteration.

### Using a `for` loop {-}

```{r}
for (i in x)
    print(log(i))
```

The loop above only prints the results on screen. They aren't stored
and are lost for any further re-use, which would be very annoying if
if took much more time to perform all the calculations. In the code
chunk below, we are going to first inititalise a vector with the
appropriate number of `NA` values and, at each iteration, we then
store the result. We however now need to change our loop and iterate
of the indices of the input vector, so that we can re-use these
indices to save the results in the output vector.


```{r}
res_loop <- rep(NA, length(x))
for (i in seq_along(x))
    res_loop[i] <- log(x[i])
res_loop
```


### Using the `apply` function {-}

The apply family of functions implements our defintion of iteration
quite literally

> Repeat `log` for each value of my input `x`

is reformulated as

> For each value of my input `x`, apply `log`

and coded as


```{r}
res_apply <- sapply(x, log)
res_apply
```

There are three such function that *apply* a function iteratively:

- `sapply` iterates over a vector and returns a new vector of the same
  length as the input vector[^sapplyexcetption].

- `lapply` iterates over a linear input (a vector of a list) and
  returns a list of the same length as the input.

- `apply` iterates of the rows or the columns of a `data.frame` or
  `matrix` and returns a list or vector or approritate length (number
  of rows or columns). The dimension over which the iterations
  proceeds is defined by the second argument, where `1` defines rows
  and `2` defined columns[^arrayapply].


[^sapplyexcetption]: This is a simplification of how `sapply` works,
    that is partly defined by the the `simplify` argument and whether
    the result of applying the function on each element of the input
    can be returned as a vector. The alternative is to return a
    `list`, like `lapply`.

[^arrayapply]: This generalises to arrays with more than 2 dimensions.


When performing the same operations in different ways (using different
implementations), it is essential to verify that the results are
identical or, if not, at least compatible. The most direct way to do
the former is to use the `identical` function:

```{r}
identical(res_loop, res_apply)
```

### Vectorisation {-}

We must not forget the obvious, which is vectorisation. Many R
functions work by default iteratively on every element of a vectors,
i.e. they work irrespectively whether the vector is of length
1[^scalar] or longer.

[^scalar]: A vector of length 1 would be called a scalar in other
    programming languages.


```{r}
res_vec <- log(x)
res_vec
```

And, as before, we check that we obtain identical results:

```{r}
identical(res_loop, res_vec)
```

### Which iteration to use?

Even though they produce the same result, the iteration strategies
above aren't equal, and some should be preferred in different
situtions.

- When a vectorised solution exists, this is the one that should be
  chosen. It is by far the fastest solution, but only applicable to
  existing functions. If you were to write your own function to
  iterate over, it is advisable to write a vectorised function.

- The apply functions are extremely convenient and concise, and hence
  widely used. They have a couple of advantages, including that there
  is no need to explicitly initialise a result variable and that they
  can easily be parallelised.

- For loops are the most generic solution for iteration, and they
  require to initialise the result variable that will be populated
  during the loop. As opposed to popular belief, they aren't slower
  than using apply functions. They are however the best solution if,
  during the iteration, one has the access another element than the
  one currently processed (typically, if `i` is the coutner, accessing
  `i + 1` or `i - 1`).




<!-- Possibly also mention `while`. -->


### Example: iterating over files
