# Conclusions {#sec-ccl}

In this course, we have seen the importance of structured data. Good
data structure starts with simple, tidy tabular data, whether it is
manually encoded in spreadsheet, or handled in R as dataframes or
tibbles. More complex data, that doesn't fit in tabular data, can be
modelled into dedicated objects that display specialised
behaviour. Structured data allows us to reason on that data, without
having to look it at. Reasoning on and generalisation of data in turn
allows to manipulate and visualise it, i.e. to explore, analyse and
understand it. The cherry on top of the data analysis cake is to be
able to reproduce an analysis, either oneself or share it in a way
that others can.

As mentioned in the preamble, the goal of this course is obviously not
for students that take it to qualify as bioinformaticians at the
end. However, what is important is to appreciate the importance of
data and their analysis, and to become fluent in exploring, discussing
and communicating around data. A shared appreciation of data and their
complexity will hopefully reduce the distinction between
bioinformaticians and experimental scientists. Indeed, at the end of
the day, it's useful to remember that

> We are all biologists, in that we study biology. Some use wet lab
> experiments, others dry lab techniques.

## Next steps

- Statistics and machine learning (see your statistics courses and the
  follow course
  [WSBIM1322](https://github.com/UCLouvain-CBIO/WSBIM1322)).
- Getting better at programming and data analysis. See
  [@r4ds:2017] and [@advancedR].
- Evolving scripts into tools/packages [@rpkgs:2015].
- Other tools: unix command line and git/GitHub [@Perez-Riverol:2016].
  See also this [short
  tutorial](https://lgatto.github.io/github-intro/).
- Omics data analysis (see upcoming
  [WSBIM2122](https://github.com/UCLouvain-CBIO/WSBIM2122) course).


## Additional exercises

To answer the following exercises, you'll need to resort to what you
have learnt in various chapters.

`r msmbstyle::question_begin()`

Make sure you have `rWSBIM1207` version >= 0.1.16 and load the 2022
Belgian road accidents statistics and the associated metadata,
describing the variables. The path to the former as an `rds` file is
available with `road_accidents_be_2022.rds()`. The
`road_accidents_be_meta.csv()` returns the path to the metadata csv
file.

The data provides the Number of killed, seriously injured, slightly
injured and uninjured victims of road accidents, by age group, type of
user, sex and various characteristics of the accident in Belgium in
20222.

- Using the appropriate functions, load both files into R and
  familiarise yourself with the data.

- Visualise the numbers for man and women over the hours of the day
  for all age classes. Ignore any unknown information. Do you see a
  difference between man and women?

- Visualise the number of victims in the different provinces. Do this
  comparison for the different type of victims. Ignore any unknown
  information. Use lines and points for this visualisation.

- Come up with additional visualisations that you could produce with
  these data. Use bar plots for this visualisation.

`r msmbstyle::question_end()`

```{r accidents, echo=FALSE, include=FALSE}
library(rWSBIM1207)
x <- readRDS(road_accidents_be_2022.rds())

## Compare the numbers for man and women over the hours of the day for
## all age classes.
dplyr::count(x, TX_SEX_DESCR_FR, DT_HOUR, TX_AGE_CLS_DESCR_FR) |>
    filter(TX_SEX_DESCR_FR != "Inconnu",
           TX_AGE_CLS_DESCR_FR != "Inconnu") |>
    ggplot(aes(x = DT_HOUR, y = n, colour = TX_SEX_DESCR_FR)) +
    geom_point() +
    geom_line() +
    facet_wrap(~ TX_AGE_CLS_DESCR_FR)


## Compare the number of victims in the different provinces. Do this
## comparison for the different type of victims.
dplyr::count(x, TX_VICT_TYPE_DESCR_FR, TX_PROV_DESCR_FR) |>
    filter(TX_VICT_TYPE_DESCR_FR != "Inconnu",
           TX_PROV_DESCR_FR != " ") |>
    ggplot(aes(x = TX_VICT_TYPE_DESCR_FR, y = n)) +
    geom_bar(stat = "identity") +
    facet_wrap(~ TX_PROV_DESCR_FR) +
    theme(axis.text.x = element_text(angle = 90, hjust = 0.5, vjust = 0.5))
```

`r msmbstyle::question_begin()`

For this exercise, make sure you have `rWSBIM1207` version >=
0.1.17. Using the `population_be.csv()` function, get the path to
`r length(rWSBIM1207::population_be.csv())` files with the population
numbers across multiple regions of Belgium up to 2023.

**Tip:** given that all files contain equivalent data (i.e. for the
same variables), you can use `read_csv()` to load all files as once
into a long table.

- How many regions have been survey over the years?

- What differences are there in terms of region?

Focusing on Belgium and the Brussels, Walloon and Flemish regions, and
from 1991 on, generate one figure (possibly with multiple facets) that
allows to answer the following questions:

- Has the population increased since 1991?

- Are there more women or men living in these regions? Have this
  changed since 1991?

- Have the changes in population been driven by men, women or both
  equally?

- What region has the biggest (lowest) population?

`r msmbstyle::question_end()`

```{r pop, echo=FALSE, include=FALSE}
library(tidyverse)
library(rWSBIM1207)

x <- read_csv(population_be.csv())


## How many regions have been survey over the years?
dplyr::count(x, annee)

## What differences are there in terms of region?
dplyr::count(x, annee) |>
   pull(n) |>
   unique()

x2023 <- filter(x, annee == 2023) |>
  select(lieu_de_residence, total) |>
  dplyr::rename("total2023" = "total")

x2018 <- filter(x, annee == 2018) |>
  select(lieu_de_residence, total) |>
  dplyr::rename("total2018" = "total")

## unique 2023
full_join(x2023, x2018) |> filter(is.na(total2018))

## unique 2018
full_join(x2023, x2018) |> filter(is.na(total2023))

## Focusing on Belgium and the Brussels, Walloon and Flemish regions, and
## from 1991 on, generate one figure (possibly with multiple facets) that
## allows to answer the following questions:

lieux <- x[[2]][1:4]

x |>
    pivot_longer(names_to = "variable",
                 values_to = "value",
                 3:5) |>
    filter(lieu_de_residence %in% lieux) |>
    filter(annee > 1990) |>
    ggplot(aes(x = annee, y = value,
               colour = variable)) +
    geom_line() +
    facet_grid(. ~ lieu_de_residence)

x |>
    pivot_longer(names_to = "variable",
                 values_to = "value",
                 3:5) |>
    filter(lieu_de_residence %in% lieux) |>
    filter(annee > 1990) |>
    ggplot(aes(x = annee, y = value,
               colour = variable)) +
    geom_line() +
    facet_wrap(~ lieu_de_residence, scale = "free_y")
```
