---
title: "Introduction to bioinformatics"
author: "Laurent Gatto"
date: "`r Sys.Date()`"
site: bookdown::bookdown_site
knit: bookdown::preview_chapter
description: "Course material for the Introduction to bioinformatics (WSBIM1207) course at UCLouvain."
output:
  msmbstyle::msmb_html_book:
    toc: TRUE
    toc_depth: 1
    split_by: chapter
    split_bib: no
    margin_references: TRUE
link-citations: yes
bibliography: [refs.bib, packages.bib]
---


```{r setup00}
knitr::knit_hooks$set(answer = function(before, options, envir) {
    if (before) {
        paste(
            "<div class=\"accordion\">",
              "<h3 class=\"toc-ignore\">Answer</h3>",
              "<div style=\"background: #fff;\">",  sep = "\n")
    } else {
        paste("</div>", "</div>", sep = "\n")
    }
})


eng_text_answer <- knitr:::eng_html_asset(
                               paste(
                                   "<div class=\"accordion\">",
                                   "<h3 class=\"toc-ignore\">Answer</h3>",
                                   "<div style=\"background: #fff;\">",
                                   "<p>",  sep = "\n"),
                               paste(
                                   "</div>", "</div>", "</p>", sep = "\n"
                               )
                           )

knitr::knit_engines$set(text_answer = eng_text_answer)
```

# Preamble {-}

The [WSBIM1207](https://uclouvain.be/cours-2018-wsbim1207.html) course
is an introduction to bioinformatics (and data science) for biology
and biomedical students. It introduces bioinformatics methodology and
technologies without relying on any prerequisites. The aim of this
course is for students to be in a position to understand important
notions of bioinformatics and tackle simple bioinformatics-related
problems in R, in particular to develope simple R analysis scripts and
reproducible analysis reports to interogate, visualise and understand
data in a tidy tabular format.

The course will be followed by *Bioinformatics*
([WSBIM1322](https://github.com/UCLouvain-CBIO/WSBIM1322)) and *Omics
data analysis*
([WSBIM2122](https://github.com/UCLouvain-CBIO/WSBIM2122)).

It is interesting to start this course by asking the students[^qst], who
have likely not yet been exposed to bioinformatics

> What is bioinformatics?

[^qst]: The same question will be asked at the very end of the
course. It will be interesting to assess how the students'
appreciation of data science and bioinformatics has evolved.

While [Wikipedia
defines](https://en.wikipedia.org/wiki/Bioinformatics) it as

> Bioinformatics is an interdisciplinary field that develops methods
  and software tools for understanding biological data.

this course will focus on the doing, rather than on knowing
definitions. Hence, the material and how it is taught will be focused
on practice.

## Motivation {-}

Although bioinformatics is playing a central role in bio-medical
research, bioinformatics skills aren't well integrated in life science
curricula, limiting students in their career prospects and research
horizon [@WilsonSayres:2018]. It is important for young researchers to
acquire quantitative, computational and data skills to address the
challenges that lie ahead.

This first course will focus on the data skills. It will not teach to
use any specific piece of bioinformatics software. Once one has
identified a relevant piece of software, running it shouldn't be a
major problem[^runningsw]. The important part will be to understand
what it does, why it does it and how to assess if the output one's
data can be trusted. The latter requires to explore and understand the
data and the results, i.e. data skills. As described by @Buffalo:2015,
equating learning a piece of bioinformatics software to learning
bioinformatics is like learning pipetting as a means to learn
molecular biology.

[^runningsw]: Although, in all fairness, some bioinformatics software
are notoriously difficult to install and run.

Critical thinking is essential in any aspect of research, and
bioinformatics and data analysis doesn't escape that rule. Teaching
critical thinking is however very difficult, and arguably impossible
without first possessing the tools to be critical. Rather that
teaching a limited set of software tools, the course aims at teaching
the skills that will allow to interrogate and visualise data, and hence
be in a position to be critical with respect to ones data. In
addition, while specific software become obsolete and are replaced, or
are specific to one specific field of research, critical investigation
of data will always be required and will never be replaced.

A final reason why the focus of the course ought to be on data skills
is that an essential difficulty in modern, multidisciplinary research
is communication. Wetlab biomedical scientists aren't required to
become bioinformaticians, statisticians, programmers, ... to be
outstanding researchers, but they will need to communicate efficiency
with these experts (and vice versa). What most often unites all these
experts is data, and communication around data is critical. The
importance of critical thinking and communication around data becomes
more evident when one realises that, when tracking work in
bioinformatics core facilities, only a minority of projects were
purely routine and that most researchers came to the bioinformatics
core seeking customized analysis, rather than a standardized package
[@Chang:2015].

## References and credits {-}

References are provided throughout the course. Several stand out
however, as they cover large parts of the material or provide
complementary resources.

The material for the first chapters, covering the *Introduction to
data science with R*, is based on the [**Data Carpentry** Ecology
curiculum](https://datacarpentry.org/lessons/#ecology-workshop).

General references for this course are *R for Data Science*
[@r4ds:2015] and *Bioinformatics Data Skills* [@Buffalo:2015].

## About this course material {-}

This material is written in R markdown [@R-rmarkdown] and compiled as a
book using `knitr` [@R-knitr] `bookdown` [@R-bookdown]. The source
code is publicly available in a Github repository
[https://github.com/uclouvain-cbio/WSBIM1207](https://github.com/uclouvain-cbio/WSBIM1207)
and the compiled material can be read at http://bit.ly/WSBIM1207.

Contributions to this material are welcome. The best way to contribute
or contact the maintainers is by means of pull requests and
issues. Please familiarise yourself with the [code of
conduct](https://github.com/UCLouvain-CBIO/WSBIM1207/blob/master/CONDUCT.md). By
participating in this project you agree to abide by its terms.


## Setup {-}

We will be using the [R environment for statistical
computing](https://www.r-project.org/) as main data science language.
We will also use the
[RStudio](https://www.rstudio.com/products/RStudio/) interface to
interact with R and write scripts and reports. Both R and RStudio are
easy to install and works on all major operating systems.

Once R and RStudio are installed, a set of packages will need to be
installed. See section \@ref(sec:setup2) for details.

To build this book, you'll need `bookdown` [@R-bookdown] and a
fork[^msmbfork] of [`msmbstyle`
style](https://github.com/grimbough/msmbstyle/) [@R-msmbstyle].

[^msmbfork]: https://github.com/lgatto/msmbstyle

```{r combilebook1, eval=FALSE}
install.packages("bookdown")
devtools::install_github("lgatto/msmbstyle")
```

In the course's work directory, simply type

```{r combilebook2, eval=FALSE}
bookdown::render_book(".")
```


## License {-}

This material is licensed under the [Creative Commons
Attribution-ShareAlike 4.0
License](https://creativecommons.org/licenses/by-sa/4.0/)/
